Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-04-20 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
draft intro + new refs

These properties are not mere theoretical curiosities; they have important observational consequences.
%
First, the presence of the scalar charge implies that when in binaries, BHs can source
scalar-dipole radiation (see, e.g., Refs.~\cite{Yagi:2011xp,Julie:2019sab,Shiralilou:2020gah,Shiralilou:2021mfl,Julie:2022huo})
which affects the GW phase at $-1$PN order (relative to the dominant quadrupolar GR
contribution), with magnitude proportional to the difference between the
charges of binary components. This makes EdGB gravity testable with present GW
observations of compact binaries where at least if one component is a BH.
%
More specifically, Ref.~\cite{Lyu:2022gdr} placed the bound
%
$\ell_{\rm EdGB} \leqslant 1.33$~km
%
using the neutron-star--black-hole binaries GW200105 and
GW200115~\cite{LIGOScientific:2021qlt} while Refs.~\cite{Nair:2019iur,Perkins:2021mhb}, obtained
%
$\ell_{\rm EdGB} \leqslant 1.7$~km,
%
by stacking the posteriors of $\ell_{\rm EdGB}$ from a selection of binary BHs (BBH)
from the GWTC-1 and GWTC-2 catalogs~\cite{LIGOScientific:2018mvr,LIGOScientific:2020ibl}~\footnote{These bounds
are, strictly speaking, valid only when the scalar field $\varphi$ is small (i.e., $\varphi \ll 1$), and
take into account only the leading-order scalar field contribution arising from the dilatonic coupling.}.
Secondly, the scalar field does influence the response of BHs to linearized perturbations and thus
affects the BH QNM spectra.
%
The coupling between scalar field and the Gauss-Bonnet invariant, which
result in a coupling between scalar perturbations and gravitational
perturbations of polar parity can, for example break the equivalence
between the QNM spectra of polar and axial gravitational perturbations~\cite{Glampedakis:2017rar,Lenzi:2021njy}
(i.e., the isospectrality~\cite{Chandrasekhar:1985kt}) for Schwarzschild BHs in GR.
%

@@

The theory admits as a solution, the garden-variety Schwarzschild BH of GR. This
is not the case when rotation is included and the Kerr metric is not a solution of
the theory~\cite{Jackiw:2003pm}.
%
These rotating BHs support a scalar field which fall off as
$r^{-2}$ asymptotically, to which we can associate a scalar dipole
charge~\cite{Yunes:2009hc,Konno:2009kg}.
%
As a consequence, the leading-scalar radiative channel of BBH to lose energy is quadrupolar,
which affects the GW phase at 2PN order~\cite{Yagi:2011xp,Alexander:2017jmt}.
%
Deviations from GR at this PN order are constrained with present GW observations~\cite{LIGOScientific:2021sio}
\ab{only at the level of $\sim 50\%$ (see the constraint on the 2PN parameter $\varphi_4$ in Fig.XXX of
Ref.~\cite{LIGOScientific:2021sio}). Thus, dCS gravity is poorly constrained by
analyses that rely on the inspiral portion of GWs when using current BBH observations are used~\cite{Nair:2019iur,Perkins:2021mhb}.}
%
Nonetheless, the theory has been constrained in Ref.~\cite{Silva:2020acr}, which found
%
$\ell_{\rm dCS} \leqslant 8.5$~km,
%
by folding data from the X-ray observations of the isolated neutron star
PSR~J0030+0451~\cite{Lommen:2000yt,NANOGrav:2017wvv} by
NICER~\cite{Riley:2019yda,Miller:2019cac} and from the GW observation of the binary
neutron star GW170817~\cite{TheLIGOScientific:2017qsa,LIGOScientific:2018cki}.

@@

A general procedure to describe deviations to the QNM frequencies $\omega^{\rm GR}_{\ell m n}$ and
damping times $\tau^{\rm GR}_{\ell m n}$ of BHs of GR is to write,
%
\begin{subequations}
\label{eq:general_deviation}
\begin{align}
\omega_{\ell m n} &= \omega_{\ell m n}^{\rm GR} \, (1 + \delta \omega_{\ell m n}), \\
\tau_{\ell m n }   &= \tau_{\ell m  n}^{\rm GR}   \, (1 + \delta \tau_{\ell m n }),
\end{align}
\end{subequations}
%
where $\delta\omega_{\ell m n}$ and $\delta\tau_{\ell m  n}$ are the fractional deformation parameters,
$\ell$ and $m$ are the multipole indices, and $n$ the overtone number.
%
This type of parametrization\footnote{\hs{See also
Refs.~\cite{Glampedakis:2017dvb,Glampedakis:2019dqh,Silva:2019scu} and
Refs.~\cite{Cardoso:2019mqo,McManus:2019ulj,Kimura:2020mrh} for alternative parametrizations.}}
was adopted, for instance, in Refs.~\cite{Gossan:2011ha,Meidam:2014jpa,Brito:2018rfr,Carullo:2018sfu,Ghosh:2021mrv}.

The current LIGO-Virgo-KAGRA tests of BH ringdown (see Sec.VII.A
of Ref.~\cite{LIGOScientific:2020tif} or Sec.VIII.A
of Ref.~\cite{LIGOScientific:2021sio}) take a restrictive theory-independent
approach towards the inference of $\delta\omega_{\ell m n}$ and
$\delta\tau_{\ell m n}$. These deviations are either assumed to occur
identically across all observed sources or belong to a generic
underlying Gaussian population. A more realistic scenario, however, is
that these parameters depend on the source BH's mass and spin.
%
Ideally, one would like to explicitly reinstate this dependence, by
%
(i) introducing deformation parameters which can be determined, once and for
all, from a specific gravity theory (GR included), and
%
(ii) making it simpler to combine constraints coming from multiple (independently observed) sources.

The ParSpec framework was introduced in Ref.~\cite{Maselli:2019mjd} and can be
used to our purpose. It is an observable-based bivariate expansion of
Eq.~\eqref{eq:general_deviation}, given by
%
\begin{subequations}
\label{eq:kerr_expansion}
\begin{align}
\omega_{\ell m n} &= \frac{1}{M_{\rm f}} \sum_{j = 0}^{N_{\rm max}} \, \chi_{\rm f}^{j} \omega^{(j)}_{\ell m n} \, \left( 1 + \gamma\, \delta \omega^{(j)}_{\ell m n} \right), \\
\tau_{\ell m n}   &= M_{\rm f}     \sum_{j = 0}^{N_{\rm max}} \, \chi_{\rm f}^{j} \tau^{(j)}_{\ell m n}   \, \left( 1 + \gamma\, \delta \tau^{(j)}_{\ell m n} \right),
\end{align}
\end{subequations}
%
where $M_{\rm f}$ and $\chi_{\rm f}$ are the detector-frame final mass and spin, respectively;
the quantities $\omega_{\ell m n}^{(j)}$ and $\tau_{\ell m n}^{(j)}$ are dimensionless coefficients of the
expansion in spin for the QNMs of BHs in GR, while $\delta\omega_{\ell m n}^{(j)}$ and $\delta\tau_{\ell m n}^{(j)}$ are
source-independent dimensionless coefficients that characterize the corrections to the GR QNM at
each spin-order,
%
and $N_{\rm max}$ is the order of the spin expansion.
%
All source dependence due to a given modified gravity theory is contained in the
dimensionless parameter $\gamma$, which reads
%
\begin{equation}
\gamma
%
= \left(\frac{\ell_{\rm th}}{M^{\rm s}_{\rm f}}\right)^{p}
%
= \left[
\frac{\ell_{\rm th} c^2 (1 + z)}{G M_{\rm f}}
\right]^{p},
\label{eq:def_gamma}
\end{equation}
%
which depends on the length-scale parameter $\ell_{\rm th}$ of the specific gravity theory
(non-GR modifications become important at distances $\lesssim \ell_{\rm th}$), and
the exponent $p$ is related to how the non-GR modifications are
added to the Einstein-Hilbert action.
%
In Eq.~\eqref{eq:def_gamma} we made $\gamma$ dimensionless by \redsout{using} the \redsout{other}
length-scale associated with the remnant black hole (i.e., its source-frame mass
$M^{\rm s}_{\rm f}$\footnote{Throughout this paper, we maintain the convention of
using the upperscript ``s" to denote source-frame quantities and plain baseline
symbols for detector-frame measurements.}), which we can also write in terms of the detector-frame mass $M_{\rm f}$
through the redshift $z$~\cite{Krolak:1987ofj}. Also, dividing by the factor $G/c^2$ allows us to
express $\ell_{\rm th}$ in physically intuitive metric units.

@@

We now describe the waveform model used in our paper to infer properties of a
BBH ringdown.
%
As in Refs.~\cite{Brito:2018rfr,Ghosh:2021mrv}, we use an IMR BBH waveform
model where the (complex\hs{-valued}) frequencies describing the remnant object
are left additionally free and estimated directly from the data.

@@

In this section, we first provide a basic outline of the Bayesian formalism that we use to
infer the properties of the underlying GW signal; then, we identify the most promising
events from the catalog of LIGO-Virgo-KAGRA GW observations to base our analyses on. Finally,
we discuss how we can interpret our results after taking into account the region of validity of
the non-GR theories that we are considering, which are EFTs.

@@

\subsection{Event selection}

@@

The \pSEOB{} model, as described in Sec.~\ref{sec:review_pSEOB}, is an
IMR model that infers the properties of the underlying GW signal,
including (independently) its ringdown properties, using the Bayesian formalism
above. Naturally, the most promising candidates for our analyses are high-mass
\emph{and} loud GW observations with a significant signal-to-noise ratio (SNR)
in the post-merger part of the signal.
%
The latest LIGO-Virgo-KAGRA GW catalog~\cite{LIGOScientific:2021djp} reported 90 observed
signals not all of which are relevant for our BH ringdown analysis. In fact, in
the accompanying paper~\cite{LIGOScientific:2021sio} on tests of GR, the \texttt{pSEOBNRv4HM}
~\cite{Brito:2018rfr,Ghosh:2021mrv} analysis~\footnote{See, in particular, Sec.VIII A.2
in Ref.~\cite{LIGOScientific:2021sio}}, which is most similar to the \pSEOB{} model
presented in this paper, identified two events which provided the strongest
bounds on the measurements of the dominant $(220)$ QNM:
GW150914~\cite{LIGOScientific:2016aoc} and
GW200129~\cite{LIGOScientific:2021djp}.
%
These two events, with a total (source-frame) mass of $65 \Mo$ and $63.4 \Mo$ respectively, are
extremely similar in their source properties. These are also two of the loudest
BBH signals observed to date with a total network SNR of 24 and 26.8,
respectively.
%
Moreover, and what is more relevant for our analysis, are their post-inspiral
(merger-ringdown) SNRs which are both $\sim `6$ (see the columns for
$\rho_{\text{post-insp}}$ in Table III of Ref.~\cite{LIGOScientific:2019fpa} and
Table IV of Ref.~\cite{LIGOScientific:2021sio}).
%
In this paper, we are going to focus on these two GW events as our probes of
the BH ringdown in modified theories of gravity.

@@

There are two conditions that we must verify before we can confidently claim to
have placed a constraint on $\ell_{\rm th}$. First, as we have explained in Sec.~\ref{sec:review_theories}, all theories that we
consider must be interpreted as an effective field theory, meaning that it should be
considered valid only below an energy scale, or equivalently, a length-scale.
%
As a cut-off length-scale for the validity of the EFT we use,
%
\begin{equation}
\Lambda_{\rm EFT} (\varepsilon, \mathfrak{m}) = \varepsilon \, \frac{G \mathfrak{m}}{c^2} \,,
\label{eq:def_cutoff}
\end{equation}
%
where $\varepsilon$ is a dimensionless number and $\mathfrak{m}$ is the median
value of one of the mass scales involved in the problem.
%
We note that $\Lambda_{\rm EFT}$ has dimensions of length and hence can be compared to the coupling strength $\ell_{\rm th}$.
%
\hs{Here we explore the range $\varepsilon \in [0, 1]$, but following
the Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr} we quote our final results using
$\varepsilon = 1/2$.}
% %
% In Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}, $\varepsilon = 1/2$ was used,
% but here we explore $\varepsilon \in [0, 1]$ for generality.
% %
We say that a bound has been placed on $\ell_{\rm th}$, if most of the PDF $P(\ell_{\rm th} | d)$ support is in the interval
$[0, \, \Lambda_{\rm EFT}(\hs{1/2}, \mathfrak{m})]$.
%
In practice, this can be quantified through the cumulative distribution function
(CDF) associated with the marginalized posterior distribution $P(\ell_{\rm th} | d)$, namely
%
\begin{equation}
P(\ell_{\rm th} \leqslant \ell^{\rm max}_{\rm th} | d) = \int_{0}^{\,\ell^{\rm max}_{\rm th}} \dd \ell'\, P(\ell' | d).
\label{eq:def_cdf}
\end{equation}
%
For instance, we require that for a bound at 90\% credible level to be placed on $\ell_{\rm th}$ that
%
\begin{equation}
% \textrm{CDF}(\ell) \leqslant \textrm{CDF}(T_{\rm EFT}(\varepsilon; \, \mathfrak{m})).
P(\ell_{\rm th} \leqslant \Lambda_{\rm EFT} | d) \leqslant 0.9,
\quad \textrm{(EFT bound)},
\label{eq:eft_bound}
\end{equation}
%
where we let $\ell^{\rm max}_{\rm th} = \Lambda_{\rm EFT}$ in Eq.~\eqref{eq:def_cdf}, and
likewise for other credibility percentiles.

@@

\begin{figure}[h]
\includegraphics[width=\columnwidth]{figs/edgb_gamma.pdf}
\caption{(Color online).~Posterior distribution function on the dimensionless parameter $\gamma_{\rm EdGB}$,
defined in Eq.~\eqref{eq:def_gamma_gb}. Different line colors distinguish between events, while
different line styles distinguish between different $N_{\rm max}$.
%
We see that this parameter which controls the ParSpec expansion in EdGB gravity
does have maximum support at $\gamma_{\rm EdGB} = 0$. This shows that our model is consistent with GR.
\hs{The right panel shows the PDFs in the range $0 \leqslant \gamma_{\rm EdGB} \leqslant 1$. Note how
the curves are flat in this range for both events and when $N_{\rm max} = 1$.}
}
\label{fig:gamma_gb}
\end{figure}

@@

Let us now consider the quartic EFT of GR, as our final example.
%
In Fig.~\ref{fig:qEFT_exec_sum} we show the posteriors on $\ell_{\rm qEFT}$ for GW150914 (top panel),
GW200129 (middle panel) for $N_{\rm max} = 0$, which are qualitatively similar to the cubic EFT of GR.
%
We find that both the EFT~\eqref{eq:eft_bound} and ParSpec~\eqref{eq:parspec_bound} are satisfied,
hence allowing us to infer bounds on $\ell_{\rm qEFT}$.
%
As Fig.~\ref{fig:qEFT_cdf} shows the CDFs for both events reach 90\% for $\gm = M_{\rm f}$ before $\varepsilon = 1/2$.
%
Consequently, from the joint posterior of $\ell_{\rm qEFT}$ we quote our final result,
%
\begin{equation}
    \ell_{\rm qEFT} \leqslant 51.3~\textrm{km} \quad \textrm{at 90\% credible level}\,.
\end{equation}
%
% \abcomm{Please update the paragraph below.}
\hs{
In this theory, we have considered only $N_{\rm max} = 0$. We found that the
inclusion of the spin corrections yields waveforms which for $\ell_{\rm qEFT} \gtrsim 80$~km
for which the ringdown portions of the larger than the inspiral-plunge part.
%
To avoid this issue we have lowered the value of $\ell^{\rm max}_{\rm qEFT}$, but by doing so we have obtained posteriors
which were essentially uniform, as our prior, and thus uninformative. Hence, we do not quote any results for $N_{\rm max} = 1$.
}
%
Table~\ref{tab:summary_qeft} summarizes our findings for the quartic EFT of GR.
%
\begin{figure}[t]
\includegraphics[width=\columnwidth]{figs/qeft_posteriors_combined.pdf}
\caption{(Color online).~Posterior distribution function for the coupling constant $\ell_{\rm qEFT}$ for
GW150914 (top panel), GW200129 (middle panel) and combined events (bottom panel).
}
\label{fig:qEFT_exec_sum}
\end{figure}

@@

We presented an unified framework to combine the ParSpec formalism introduced
to model deviations to the GR QNMs~\cite{Maselli:2019mjd} with the \pSEOB~waveform model~\cite{Brito:2018rfr,Ghosh:2021mrv}.
%
We showed with concrete examples, how theory-specific QNM calculations of
slowly-rotating BHs in modified gravity theories can be mapped onto the
non-GR parameters of the ParSpec formalism.
%
Put together this allowed us to use the \pSEOB~waveform model to test four
modified gravity theories (EdGB, dCS, cubic, and quartic EFTs of GR) using
observational data from the LIGO-Virgo-KAGRA's events GW150914 and GW200129.
%
We found, in particular that, within the interpretation of our non-GR theories as EFTs and the
region of validity of the ParSpec framework, the coupling constant of dCS gravity is bound to
$\ell_{\rm CS} \leqslant 34.5$~km, at 90\% credible level, when stacking the posteriors
of GW150914 and GW200129.
%
In contrast, we could not place any constraint of the coupling constants in EdGB gravity $\ell_{\rm GB}$.
%
This dichotomy between these theories has an analogy with works that considered
the inspiral part of the GW signal alone.
%
In that case, it is dCS which is unconstrained \hs{with already observed GW
events}, while EdGB can be constrained, as shown in Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}.
%
\hscomm{I will a sentence here on Refs.~\cite{Alexander:2017jmt,Loutrel:2018ydv,Loutrel:2018rxs}.}

@@

\hscomm{TODO: add that we could try to parametrize the final black hole mass and spin estimates formulas we
use at the moment~\cite{Hofmann:2016yih}. Maybe comment that a possible first attempt could follow~\cite{Buonanno:2007sv},
as done in~\cite{Jai-akson:2017ldo,Li:2020wse}. Say that ultimately we will need to informed by NR in beyond-GR.}

>>>>

These properties are not mere theoretical curiosities; they have important observational consequences.
%
First, the presence of the scalar charge implies that when in binaries, BHs can source
scalar-dipole radiation (see, e.g., Refs.~\cite{Yagi:2011xp,Julie:2019sab,Shiralilou:2020gah,Shiralilou:2021mfl,Julie:2022huo})
which affects the GW phase at $-1$PN order (relative to the dominant quadrupolar GR
contribution), with magnitude proportional to the difference between the
charges of binary components. This makes EdGB gravity testable with present GW
observations of compact binaries where at least if one component is a BH.
%
More specifically, Ref.~\cite{Lyu:2022gdr} placed the bound
%
$\ell_{\rm EdGB} \leqslant 1.33$~km
%
using the NSBH binaries GW200105 and
GW200115~\cite{LIGOScientific:2021qlt} while Refs.~\cite{Nair:2019iur,Perkins:2021mhb}, obtained
%
$\ell_{\rm EdGB} \leqslant 1.7$~km,
%
by stacking the posteriors of $\ell_{\rm EdGB}$ from a selection of BBHs 
from the GWTC-1 and GWTC-2 catalogs~\cite{LIGOScientific:2018mvr,LIGOScientific:2020ibl}~\footnote{These bounds
are, strictly speaking, valid only when the scalar field $\varphi$ is small (i.e., $\varphi \ll 1$), and
take into account only the leading-order scalar field contribution arising from the dilatonic coupling.}.
Secondly, the scalar field does influence the response of BHs to linearized perturbations and thus
affects the BH QNM spectra.
%
The coupling between scalar field and the Gauss-Bonnet invariant, which
result in a coupling between scalar perturbations and gravitational
perturbations of polar parity can, for example break the equivalence
between the QNM spectra of polar and axial gravitational perturbations~\cite{Glampedakis:2017rar,Lenzi:2021njy}
(i.e., the isospectrality~\cite{Chandrasekhar:1985kt}) for Schwarzschild BHs in GR.
%

@@

The theory admits as a solution, the garden-variety Schwarzschild BH of GR. This
is not the case when rotation is included and the Kerr metric is not a solution of
the theory~\cite{Jackiw:2003pm}.
%
These rotating BHs support a scalar field which fall off as
$r^{-2}$ asymptotically, to which we can associate a scalar dipole
charge~\cite{Yunes:2009hc,Konno:2009kg}.
%
As a consequence, the leading-scalar radiative channel of BBH to lose energy is quadrupolar,
which affects the GW phase at 2PN order~\cite{Yagi:2011xp,Alexander:2017jmt}.
%
Deviations from GR at this PN order are constrained with present GW observations~\cite{LIGOScientific:2021sio}
\ab{only at the level of $\sim 50\%$ (see the constraint on the 2PN parameter $\varphi_4$ in Fig.XXX of
Ref.~\cite{LIGOScientific:2021sio}). Thus, dCS gravity is poorly constrained by
analyses that rely on the inspiral portion of GWs when using current BBH observations are used~\cite{Nair:2019iur,Perkins:2021mhb}.}
%
Nonetheless, the theory has been constrained in Ref.~\cite{Silva:2020acr}, which found
%
$\ell_{\rm dCS} \leqslant 8.5$~km,
%
by folding data from the X-ray observations of the isolated NS 
PSR~J0030+0451~\cite{Lommen:2000yt,NANOGrav:2017wvv} by
NICER~\cite{Riley:2019yda,Miller:2019cac} and from the GW observation of the binary
NS GW170817~\cite{TheLIGOScientific:2017qsa,LIGOScientific:2018cki}.

@@

A general procedure to describe deviations to the QNM frequencies $\omega^{\rm GR}_{\ell m n}$ and
damping times $\tau^{\rm GR}_{\ell m n}$ of BHs of GR is to write,
%
\begin{subequations}
\label{eq:general_deviation}
\begin{align}
\omega_{\ell m n} &= \omega_{\ell m n}^{\rm GR} \, (1 + \delta \omega_{\ell m n}), \\
\tau_{\ell m n }   &= \tau_{\ell m  n}^{\rm GR}   \, (1 + \delta \tau_{\ell m n }),
\end{align}
\end{subequations}
%
where $\delta\omega_{\ell m n}$ and $\delta\tau_{\ell m  n}$ are the fractional deformation parameters,
$\ell$ and $m$ are the multipole indices, and $n$ the overtone number.
%
This type of parametrization~\footnote{See also
Refs.~\cite{Glampedakis:2017dvb,Glampedakis:2019dqh,Silva:2019scu} and
Refs.~\cite{Cardoso:2019mqo,McManus:2019ulj,Kimura:2020mrh} for alternative parametrizations.}
was adopted, for instance, in Refs.~\cite{Gossan:2011ha,Meidam:2014jpa,Brito:2018rfr,Carullo:2018sfu,Ghosh:2021mrv}.

The current LVK tests of BH ringdown (see Sec.VII.A
of Ref.~\cite{LIGOScientific:2020tif} or Sec.VIII.A
of Ref.~\cite{LIGOScientific:2021sio}) take a restrictive theory-independent
approach towards the inference of $\delta\omega_{\ell m n}$ and
$\delta\tau_{\ell m n}$. These deviations are either assumed to occur
identically across all observed sources or belong to a generic
underlying Gaussian population. A more realistic scenario, however, is
that these parameters depend on the source BH's mass and spin.
%
Ideally, one would like to explicitly reinstate this dependence, by
%
(i) introducing deformation parameters which can be determined, once and for
all, from a specific gravity theory (GR included), and
%
(ii) making it simpler to combine constraints coming from multiple (independently observed) sources.

The ParSpec framework was introduced in Ref.~\cite{Maselli:2019mjd} and can be
used to our purpose. It is an observable-based bivariate expansion of
Eq.~\eqref{eq:general_deviation}, given by
%
\begin{subequations}
\label{eq:kerr_expansion}
\begin{align}
\omega_{\ell m n} &= \frac{1}{M_{\rm f}} \sum_{j = 0}^{N_{\rm max}} \, \chi_{\rm f}^{j} \omega^{(j)}_{\ell m n} \, \left( 1 + \gamma\, \delta \omega^{(j)}_{\ell m n} \right), \\
\tau_{\ell m n}   &= M_{\rm f}     \sum_{j = 0}^{N_{\rm max}} \, \chi_{\rm f}^{j} \tau^{(j)}_{\ell m n}   \, \left( 1 + \gamma\, \delta \tau^{(j)}_{\ell m n} \right),
\end{align}
\end{subequations}
%
where $M_{\rm f}$ and $\chi_{\rm f}$ are the detector-frame final mass and spin, respectively;
the quantities $\omega_{\ell m n}^{(j)}$ and $\tau_{\ell m n}^{(j)}$ are dimensionless coefficients of the
expansion in spin for the QNMs of BHs in GR, while $\delta\omega_{\ell m n}^{(j)}$ and $\delta\tau_{\ell m n}^{(j)}$ are
source-independent dimensionless coefficients that characterize the corrections to the GR QNM at
each spin-order,
%
and $N_{\rm max}$ is the order of the spin expansion.
%
All source dependence due to a given modified gravity theory is contained in the
dimensionless parameter $\gamma$, which reads
%
\begin{equation}
\gamma
%
= \left(\frac{\ell_{\rm th}}{M^{\rm s}_{\rm f}}\right)^{p}
%
= \left[
\frac{\ell_{\rm th} c^2 (1 + z)}{G M_{\rm f}}
\right]^{p},
\label{eq:def_gamma}
\end{equation}
%
which depends on the length-scale parameter $\ell_{\rm th}$ of the specific gravity theory
(non-GR modifications become important at distances $\lesssim \ell_{\rm th}$), and
the exponent $p$ is related to how the non-GR modifications are
added to the Einstein-Hilbert action.
%
In Eq.~\eqref{eq:def_gamma} we made $\gamma$ dimensionless by \redsout{using} the \redsout{other}
length-scale associated with the remnant BH (i.e., its source-frame mass
$M^{\rm s}_{\rm f}$\footnote{Throughout this paper, we maintain the convention of
using the upperscript ``s" to denote source-frame quantities and plain baseline
symbols for detector-frame measurements.}), which we can also write in terms of the detector-frame mass $M_{\rm f}$
through the redshift $z$~\cite{Krolak:1987ofj}. Also, dividing by the factor $G/c^2$ allows us to
express $\ell_{\rm th}$ in physically intuitive metric units.

@@

We now describe the waveform model used in our paper to infer properties of a
BBH ringdown.
%
As in Refs.~\cite{Brito:2018rfr,Ghosh:2021mrv}, we use an IMR BBH waveform
model where the (complex-valued) frequencies describing the remnant object
are left additionally free and estimated directly from the data.

@@

In this section, we first provide a basic outline of the Bayesian formalism that we use to
infer the properties of the underlying GW signal; then, we identify the most promising
events from the catalog of LVK GW observations to base our analyses on. Finally,
we discuss how we can interpret our results after taking into account the region of validity of
the non-GR theories that we are considering, which are EFTs.

@@

\subsection{Events selection}

@@

The \pSEOB{} model, as described in Sec.~\ref{sec:review_pSEOB}, is an
IMR model that infers the properties of the underlying GW signal,
including (independently) its ringdown properties, using the Bayesian formalism
above. Naturally, the most promising candidates for our analyses are high-mass
\emph{and} loud GW observations with a significant signal-to-noise ratio (SNR)
in the post-merger part of the signal.
%
The latest LVK GW catalog~\cite{LIGOScientific:2021djp} reported 90 observed
signals not all of which are relevant for our BH ringdown analysis. In fact, in
the accompanying paper~\cite{LIGOScientific:2021sio} on tests of GR, the \texttt{pSEOBNRv4HM}
~\cite{Brito:2018rfr,Ghosh:2021mrv} analysis~\footnote{See, in particular, Sec.VIII A.2
in Ref.~\cite{LIGOScientific:2021sio}}, which is most similar to the \pSEOB{} model
presented in this paper, identified two events which provided the strongest
bounds on the measurements of the dominant $(220)$ QNM:
GW150914~\cite{LIGOScientific:2016aoc} and
GW200129~\cite{LIGOScientific:2021djp}.
%
These two events, with a total (source-frame) mass of $65 \Mo$ and $63.4 \Mo$ respectively, are
extremely similar in their source properties. These are also two of the loudest
BBH signals observed to date with a total network SNR of 24 and 26.8,
respectively.
%
Moreover, and what is more relevant for our analysis, are their post-inspiral
(merger-ringdown) SNRs which are both $\sim `6$ (see the columns for
$\rho_{\text{post-insp}}$ in Table III of Ref.~\cite{LIGOScientific:2019fpa} and
Table IV of Ref.~\cite{LIGOScientific:2021sio}).
%
In this paper, we are going to focus on these two GW events as our probes of
the BH ringdown in modified theories of gravity.

@@

There are two conditions that we must verify before we can confidently claim to
have placed a constraint on $\ell_{\rm th}$. First, as we have explained in Sec.~\ref{sec:review_theories}, all theories that we
consider must be interpreted as an effective field theory, meaning that it should be
considered valid only below an energy scale, or equivalently, a length-scale.
%
As a cut-off length-scale for the validity of the EFT we use,
%
\begin{equation}
\Lambda_{\rm EFT} (\varepsilon, \mathfrak{m}) = \varepsilon \, \frac{G \mathfrak{m}}{c^2} \,,
\label{eq:def_cutoff}
\end{equation}
%
where $\varepsilon$ is a dimensionless number and $\mathfrak{m}$ is the median
value of one of the mass scales involved in the problem.
%
We note that $\Lambda_{\rm EFT}$ has dimensions of length and hence can be compared to the coupling strength $\ell_{\rm th}$.
%
Here we explore the range $\varepsilon \in [0, 1]$, but following
the Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr} we quote our final results using
$\varepsilon = 1/2$. 
% %
% In Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}, $\varepsilon = 1/2$ was used,
% but here we explore $\varepsilon \in [0, 1]$ for generality.
% %
We say that a bound has been placed on $\ell_{\rm th}$, if most of the PDF $P(\ell_{\rm th} | d)$ support is in the interval
$[0, \, \Lambda_{\rm EFT}(1/2, \mathfrak{m})]$.
%
In practice, this can be quantified through the cumulative distribution function
(CDF) associated with the marginalized posterior distribution $P(\ell_{\rm th} | d)$, namely
%
\begin{equation}
P(\ell_{\rm th} \leqslant \ell^{\rm max}_{\rm th} | d) = \int_{0}^{\,\ell^{\rm max}_{\rm th}} \dd \ell'\, P(\ell' | d).
\label{eq:def_cdf}
\end{equation}
%
For instance, we require that for a bound at 90\% credible level to be placed on $\ell_{\rm th}$ that
%
\begin{equation}
% \textrm{CDF}(\ell) \leqslant \textrm{CDF}(T_{\rm EFT}(\varepsilon; \, \mathfrak{m})).
P(\ell_{\rm th} \leqslant \Lambda_{\rm EFT} | d) \leqslant 0.9,
\quad \textrm{(EFT bound)},
\label{eq:eft_bound}
\end{equation}
%
where we let $\ell^{\rm max}_{\rm th} = \Lambda_{\rm EFT}$ in Eq.~\eqref{eq:def_cdf}, and
likewise for other credibility percentiles.

@@

\begin{figure}[h]
\includegraphics[width=\columnwidth]{figs/edgb_gamma.pdf}
\caption{(Color online).~Posterior distribution function on the dimensionless parameter $\gamma_{\rm EdGB}$,
defined in Eq.~\eqref{eq:def_gamma_gb}. Different line colors distinguish between events, while
different line styles distinguish between different $N_{\rm max}$.
%
We see that this parameter which controls the ParSpec expansion in EdGB gravity
does have maximum support at $\gamma_{\rm EdGB} = 0$. This shows that our model is consistent with GR. 
The right panel shows the PDFs in the range $0 \leqslant \gamma_{\rm EdGB} \leqslant 1$. Note how
the curves are flat in this range for both events and when $N_{\rm max} = 1$.}
\label{fig:gamma_gb}
\end{figure}

@@

Let us now consider the quartic EFT of GR, as our final example.
%
In Fig.~\ref{fig:qEFT_exec_sum} we show the posteriors on $\ell_{\rm qEFT}$ for GW150914 (top panel),
GW200129 (middle panel) for $N_{\rm max} = 0$, which are qualitatively similar to the cubic EFT of GR.
%
We find that both the EFT~\eqref{eq:eft_bound} and ParSpec~\eqref{eq:parspec_bound} are satisfied,
hence allowing us to infer bounds on $\ell_{\rm qEFT}$.
%
As Fig.~\ref{fig:qEFT_cdf} shows the CDFs for both events reach 90\% for $\gm = M_{\rm f}$ before $\varepsilon = 1/2$.
%
Consequently, from the joint posterior of $\ell_{\rm qEFT}$ we quote our final result,
%
\begin{equation}
    \ell_{\rm qEFT} \leqslant 51.3~\textrm{km} \quad \textrm{at 90\% credible level}\,.
\end{equation}
%
% \abcomm{Please update the paragraph below.}
In this theory, we have considered only $N_{\rm max} = 0$. We find that the
inclusion of the spin corrections yields waveforms that for $\ell_{\rm qEFT} \gtrsim 80$~km, 
for which the ringdown portions of the larger than the inspiral-plunge part. \abcomm{The last sentence 
has some problems in English. I do not udnerstand it.}
%
To avoid this issue we have lowered the value of $\ell^{\rm max}_{\rm qEFT}$, but by doing so we have obtained posteriors
which were essentially uniform, as our prior, and thus uninformative. Hence, we do not quote any results for $N_{\rm max} = 1$.
%
Table~\ref{tab:summary_qeft} summarizes our findings for the quartic EFT of GR.
%
\begin{figure}[t]
\includegraphics[width=\columnwidth]{figs/qeft_posteriors_combined.pdf}
\caption{(Color online).~Posterior distribution function for the coupling constant $\ell_{\rm qEFT}$ for
GW150914 (top panel), GW200129 (middle panel) and combined events (bottom panel).
}
\label{fig:qEFT_exec_sum}
\end{figure}

@@

We presented an unified framework to combine the ParSpec formalism introduced
to model deviations to the GR QNMs~\cite{Maselli:2019mjd} with the \pSEOB~waveform model~\cite{Brito:2018rfr,Ghosh:2021mrv}.
%
We showed with concrete examples, how theory-specific QNM calculations of
slowly-rotating BHs in modified gravity theories can be mapped onto the
non-GR parameters of the ParSpec formalism.
%
Put together this allowed us to use the \pSEOB~waveform model to test four
modified gravity theories (EdGB, dCS, cubic, and quartic EFTs of GR) using
observational data from the LVK events GW150914 and GW200129.
%
We found, in particular that, within the interpretation of our non-GR theories as EFTs and the
region of validity of the ParSpec framework, the coupling constant of dCS gravity is bound to
$\ell_{\rm CS} \leqslant 34.5$~km, at 90\% credible level, when stacking the posteriors
of GW150914 and GW200129.
%
In contrast, we could not place any constraint of the coupling constants in EdGB gravity $\ell_{\rm GB}$.
%
This dichotomy between these theories has an analogy with works that considered
the inspiral part of the GW signal alone.
%
In that case, it is dCS which is \ab{poorly constrained with current observed GW
events}, while EdGB can be constrained, as shown in Refs.~\cite{Nair:2019iur,Perkins:2021mhb,Lyu:2022gdr}.
%
\hscomm{I will add a sentence here on Refs.~\cite{Alexander:2017jmt,Loutrel:2018ydv,Loutrel:2018rxs}.}

@@

\hscomm{TODO: add that we could try to parametrize the final BH mass and spin estimates formulas we
use at the moment~\cite{Hofmann:2016yih}. Maybe comment that a possible first attempt could follow~\cite{Buonanno:2007sv},
as done in~\cite{Jai-akson:2017ldo,Li:2020wse}. Say that ultimately we will need to informed by NR in beyond-GR.}
